Rising from the Ashes: How the Milky Way Got Its Scars

### 2024-04-21 paper/ms.tex
Write new methods, move old text to appendix

\section{Methods}\label{sec:methods}

We assume the CGM of the Milky Way and GSE can be modelled using a $\beta$ profile,
\begin{equation}
  \rho(r) = \rho_0 \left( 1 + \left(\frac{r}{r_c}\right)^2 \right)^{-\frac{3}{2}\beta}\textrm{.}
\end{equation}
However, a $\beta$-profile has been shown to be a reasonable fit to the density
profile of galaxy clusters and of the $L_{*}$ galaxy NGC 3221. Whether or not
such a profile extends to lower mass galaxies at $z~2$ is unclear, but to our
knowledge there is no better motivation for another profile.

@@

\section{Modelling the Milky Way and GSE at $z=2$}

Our merger simulations are built upon reasonable models of both the Milky Way
and GSE at $z\sim2$. Much is uncertain about the properties of the Milky Way and
GSE at these redshifts.  We describe the motivation for the choices made in
constructing these models.

>>>>

\section{Methods}\label{sec:methods}
\subsection{Simulation Setup}\label{ssec:setup}
Our goal in setting up the simulations in this work is not an attempt at matching any Milky Way observable in any detail. Our aim is to construct a setup resembling the merger between the Milky Way and GSE at $z\sim2$. However, there are great uncertainities associated with the state of the Milky Way and of GSE at this time. In Appendix~\ref{app:ics}, we discuss in more detail the choices made in tuning our setup to arrive at a system which is reasonably useful for understanding the Galaxy.

Here, we provide an overview of some salient details of our setup. The interested reader is referred to Appendix~\ref{app:ics} for more details. In isolation, each galaxy is a compound halo setup, with a Hernquist dark matter halo and a gaseous halo with a $\beta$-profile. The dark matter halo is initialized to be in gravitational equilibrium with the total potential. The gaseous halo is in gravito-hydrostatic equilibrium, where the temperature is allowed to vary as a function of radius. The azimuthal velocity of the gaseous halo is given as a fraction of the circular velocity. There is no initial stellar disk or bulge, and the gaseous halo is initially metal-free. All star particles and metals are formed self-consistently. Both the central and satellite galaxies are setup in this way.

In order to combine the galaxies, we follow \citet{2021ApJ...923...92N}, and place the satellite galaxy on a retrograde orbit. In the fiducial simulation, the satellite is placed at the virial radius with the virial velocity with a circularity of $0.5$. In order to test minor changes to the orbit, we ran a grid of simulations with $\pm10\%$ in each the starting radius and velocity, and $\pm0.1$ in the circularity, for a total of $27$ simulations.

\section{Results}\label{sec:results}
\subsection{Abundance Distribution}
In Figure~\ref{fig:fig1}, briefly discussed in the Introduction, we show the abundance distribution of the Milky Way as well as two of our idealized merger simulations. A number of our idealized simulations exhibit either a bimodal or unimodal abundance distribution, and so we have selected two which are representative examples of

@@

\section{Initial Conditions}\label{app:ics}
\subsection{Update to Initial Conditions Code}
The routines to generate the initial conditions of the central and satellite galaxies are descended from the \texttt{MakeNewDisk} code first described in \tocite. This code originally allowed for a Hernquist halo, exponential gas and stellar disk, and Hernquist bulge to be constructed in equilibrium by approximating the distribution function as a Gaussian. This code was recently extended by \citet{2023MNRAS.tmp.2070B} to also allow for the construction of a gaseous halo with a $\beta$-profile which surrounds the disk. This halo is given a rotational velocity which is a fixed ratio of the circular velocity. The interested

We have made a few modifications to the \citet{2023MNRAS.tmp.2070B} code. First, we have added the ability to vary the value of $\beta$, which was previously fixed to $2/3$. Second, we can now vary the fraction of the gaseous halo's rotational velocity to the circular velocity, which was fixed at $0.4$. The central density of the $\beta$ profile is also chosen differently. We now fix the total mass inside of $R_{200}$. The user now specifies the baryon fraction within $R_{200}$ (nominally the universal baryon fraction), from which the dark matter halo mass and central density are derived. For other details of the setup, we refer the interested reader to \citet{2023MNRAS.tmp.2070B}.

A detailed description of our setup is described in Appendix~\ref{app:ics}. Briefly, for both the central and satellite galaxies we initialize a dual system of both a Hernquist halo as well as a $\beta$-profile. \red{fix: this is confusing}. The haloes are in gravito-hydrostatic equilibrium with themselved. We do not initialize either galaxy with a disk, rather allowing the disk to form self-consistently out of the collapsing profile. This is an attractive setup because it means that the disk-CGM interface as well as the abundance of star particles are formed self-consistently.

For the central galaxy, meant to imitate the Milky Way, we choose a halo mass of $5\times10^{11}\,\Msun$.\footnote{When discussing halo masses in the \texttt{MakeNewDisk} lineage, it is important to remember that a Hernquist profile is chosen such that its profile within the scale radius matches an NFW halo\red{cite} of a certain mass, size, and concentration. This means that the actual mass contained within $R_{200}$ is less than would be contained within an NFW halo, since the Hernquist halo's density drops more quickly than NFW past the scale radius. This is unimportant for studies of the internal structure of galaxies, but in our work this distinction would mean that the given orbits would need to be adjusted.} This is roughly the expected halo mass of the Milky Way at $z\sim2$ from abundance matching\red{cite} in combination with the expected stellar mass-halo mass relation \red{cite}. We allow $\beta$ to be $0.8$ and set the central density to $\sim4.1\times10^{-4}\,\Msun/\pc^3$, such that the baryon fraction within $R_{200}$ is $8\%$. The core radius is set to be $9\,\kpc$, though the resultant galaxy is not very sensitive to the core radius.

For the satellite galaxy, meant to imitate GSE, we choose a halo mass of $\sim2.2\times10^{11}\,\Msun$. We also use a $\beta$ value of $0.8$ and set the central density to $\sim3.5\times10^{-4}\,\Msun/\pc^3$, such that the baryon fraction within $R_{200}$ is $6\%$. The core radius is set to $6.5\,\kpc$, which similar to the central does not have a large impact.

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{mass_size.pdf}
  \caption{The mass and size evolution of the central (Milky Way, blue) and satellite (GSE, orange) galaxies. The mass is taken to be the stellar mass within twice the half-mass radius, and the size is taken to be the half-mass radius. In the upper panel, we also show as a horizontal line the mass of the Milky Way and GSE from the best-fit model of \citet{2021ApJ...923...92N}. This comparison is taken to be made at $3\,\Gyr$ (vertical dashed line), our proxy for $z\sim2$. A precise match is not attempted given the wide ranging uncertainties.}
  \label{fig:mass_size}
\end{figure}

Our setup is not meant to be taken literally as a representation of the evolution of the Universe from $z=\inf$ to $z\sim2$. In particular, we have made manual adjustments to the setup parameters as well as the feedback model in order to arrive at a well-behaved disk after $\sim3\,\Gyr$ of evolution. We experimented with stronger feedback models, but found that higher baryon fractions were required to achieve reasonable stellar masses. This may come from the unrealistic aspects of our setup. In particular, our potential is initially deep, while in the real universe the potential will deepen gradually over time. Thus, a galaxy can form and become self-regulating in a less intense environment than in our setup. We account for this by lowering the baryon fraction to well below the universal value.

\subsection{Orbital Parameters}\label{ssec:orbits}
In this work, we explore a very narrow region of the orbital parameter space, building on the maximum likelihood model from \citet{2021ApJ...923...92N}. We situate the satellite galaxy at a given starting radius $R_0$. Given the symmetry, the satellite is placed on the $x$-axis.

We then assign a velocity of magnitude $V_0$ with a circularity $\eta$ where $\eta=0$ means $V_0$ is directed radially inward and $\eta=1$ means the satellite is on a circular orbit. The radal velocity is set to be $\sqrt{1-\eta^2}V_0$ and the azimuthal and vertical velocities are each set to $V_0 \eta / \sqrt{2}$. The sign of the azimuthal velocity is chosen to be retrograde.

After assigning the initial position and velocity of the satellite, we rotate the satellite in the $x$-$z$ plane such that the center of mass is inclined $15\degree$ above the plane and the halo is counterrotating with respect to the central.

For the particular values of the orbital parameters, we take the fiducial model from \citet{2021ApJ...923...92N}. We set $R_0$ to be $R_{200}$, $V_0$ to be $V_{200}$ (the circular velocity at $R_{2002}$), and the circularity $\eta$ to be $0.5$. We explored a grid of values in which we varied $R_0$ and $V_0$ each by $\pm10\%$, and $\eta$ by $\pm0.1$, for a total of $27$ simulations. In this work, we focus on the outcome of two of these simulations in order to demonstrate that our main argument is, in principle, valid. A broader exploration and expansion of the orbital grid is a topic of further investigation.

We assume the CGM of the Milky Way and GSE can be modelled using a $\beta$ profile,

However, a $\beta$-profile has been shown to be a reasonable fit to the density
profile of galaxy clusters and of the $L_{*}$ galaxy NGC 3221. Whether or not
such a profile extends to lower mass galaxies at $z~2$ is unclear, but to our
knowledge there is no better motivation for another profile.
